---
title: "LAO analysis, prediction and validation"
author: "Francesco Delogu"
date: as.Date(now())
output: html_document
---

# Analysis of LAO, prediction and validation of future samples

## Set the environment

```{r setup}

options(stringsAsFactors = F, gsubfn.engine = "R")

Sys.setenv("LANGUAGE"="En")

library(tidyverse)
library(matrixStats)
library(ggridges)
library(compositions)
library(MASS)
library(caret)
library(lubridate)
library(imputeTS)
library(parallel)
library(tseries)
library(forecast)
library(olsrr)
library(fable)
library(viridis)
library(TidyMultiqc)
library(GSEABase)
library(AnnotationDbi)
library(ggnewscale)
library(WGCNA)
library(GOstats)
library(lmtest)
library(ade4)
library(rEDM)

wd <- getwd()

```


## Load data

### Load counts

```{r load counts}

load(paste0(wd, "/preprocessed/preprocessed_mats_MRS_sampels.RData"))

```

## Time dependency analysis

### Tests for time dependency

Time dependency is tested using four sattistical tets:
- Box test [G.M Ljung, G.E.P. Box, Biometrika (1978)]
- Augmented Dickey–Fuller Test [A. Banerjee, et al., Cointegration, Error Correction, and the Econometric Analysis of Non-Stationary Data (1993)]
- KPSS Test for Stationarity with null hypothesys "Level" [D. Kwiatkowski, et al., J. Econom. (1992)]
- KPSS Test for Stationarity with null hypothesys "Trend" [D. Kwiatkowski, et al., J. Econom. (1992)]


```{r autocorrelation, echo=F, message=F, warning=F}

impute_weekly <- function(mat, dates_arr){

  X <- as.character(floor(difftime(dates_arr, dates_arr[1], units="weeks")))
  colnames(mat) <- X
  filled_X <- as.character(0:X[length(X)])
  
  new_mat <- cbind(matrix(ncol=sum(!filled_X%in%X), nrow=nrow(mat)) %>%
    `colnames<-`(filled_X[!filled_X%in%X]),
    mat)[,filled_X]
  
  patched_Y <- data.frame(apply(as.data.frame(new_mat), function(x) approxfun(seq_along(x), x)(seq_along(x)), MARGIN=1))
  return(t(patched_Y))
  
}

non_stationary_EGs <- function(mat, p_val=0.01, th=1){
  # p_val = p value for the statistical tests
  # th = how many tests (out of the four implemented) have to be passed to consider the EG non-stationary
  
  #mat <- t(mat)
  stat.tests <- matrix(NA, ncol=4, nrow=ncol(mat))
  rownames(stat.tests) <- colnames(mat)
  colnames(stat.tests) <- c("LB", "ADF", "KPSS.trend", "KPSS.level")
  
  for(i in 1:ncol(mat)){
    y_tmp <- unlist(mat[,i])
    stat.tests[i,] <- c(Box.test(y_tmp, lag=1, type="Ljung-Box")$p.value,
                        adf.test(y_tmp)$p.value,
                        kpss.test(y_tmp, null=c("Trend"))$p.value,
                        kpss.test(y_tmp, null=c("Level"))$p.value)
  }
  
  stat.tests[,c(1,3,4)] <- ifelse(stat.tests[,c(1,3,4)]<p_val, 1, 0)
  stat.tests[,2] <- ifelse(stat.tests[,2]>p_val, 1, 0)
  corrplot::corrplot(stat.tests)
  
  return((1:nrow(stat.tests))[rowSums(stat.tests)>=th])

}

```


### Non-stationary EGs

```{r EG test data plots}

omic_type <- c("MG", "MT", "MP")
fun_lvl <- paste("fun", c("C_name", "D_name"), sep=".")
tax_lvl <- paste("tax", c("phylum", "class", "order", "family", "genus", "species"), sep=".")
subsets <- c(fun_lvl, tax_lvl)
set_type <- c((expand.grid(a=omic_type, b=subsets) %>% mutate(d=paste(a, b, sep=".")))$d, "MG.raw", "MT.raw", "MP.raw")

options(warn = -1)

for(i in set_type){
  if(startsWith(i, "MO")){
    cycle_mat <- paste0(i, ".", "train.svd")
  } else {
    cycle_mat <- paste0(i, ".", "train.svd.corr")
  }
  
  weekly_imputed <- t(impute_weekly(t(get(cycle_mat)$v), colnames(MT.raw.train.clr.corr)))
  EGs_cycle <- non_stationary_EGs(weekly_imputed, th=2)
  assign(paste0(i, ".NonStatEG.corr"), EGs_cycle)
  
}

options(warn = 0)

```


### Non-stationary EGs per set

```{r EG test data plots}

omic_type <- c("MG", "MT", "MP")
fun_lvl <- paste("fun", c("C_name", "D_name"), sep=".")
tax_lvl <- paste("tax", c("phylum", "class", "order", "family", "genus", "species"), sep=".")
subsets <- c(fun_lvl, tax_lvl)
set_type <- c((expand.grid(a=omic_type, b=subsets) %>% mutate(d=paste(a, b, sep=".")))$d, "MG.raw", "MT.raw", "MP.raw")

nEGs <- matrix(0, ncol=length(omic_type), nrow=length(set_type)/length(omic_type)) %>%
  `colnames<-`(omic_type) %>%
  `rownames<-`(c("raw", rev(subsets)))

for(i in omic_type){
  for(j in c("raw", subsets)){
    if(i=="MO"){
      set_cycle <- paste(i, j, "NonStatEG", sep=".")
    } else {
      set_cycle <- paste(i, j, "NonStatEG.corr", sep=".")
    }
    nEGs[j,i] <- length(get(set_cycle))
  }
}


percEGs <- matrix(0, ncol=length(omic_type), nrow=length(set_type)/length(omic_type)) %>%
  `colnames<-`(omic_type) %>%
  `rownames<-`(c("raw", rev(subsets)))

for(i in omic_type){
  for(j in c("raw", subsets)){
    if(startsWith(i, "MO")){
      set_cycle <- paste(i, j, "NonStatEG", sep=".")
      EGval_cycle <- paste(i, j, "train.svd", sep=".")
    } else {
      set_cycle <- paste(i, j, "NonStatEG.corr", sep=".")
      EGval_cycle <- paste(i, j, "train.svd.corr", sep=".")
    }
    tmp.mat <- get(EGval_cycle)$d
    N <- length(tmp.mat)
    percEGs[j,i] <- sum((tmp.mat[2:N]/sum(tmp.mat[2:N]))[(get(set_cycle))-1])
  }
}

nEGs
percEGs

short_names_NonStatES <- c("ORFs",
                           "Tax Phylum", "Tax Class", "Tax Order", "Tax Family", "Tax Genus", "Tax Species",
                           "Fun KO", "Fun Pathway")
names(short_names_NonStatES) <- c("raw",
                                        "tax.phylum", "tax.class", "tax.order", "tax.family", "tax.genus", "tax.species",
                                        "fun.D_name", "fun.C_name")

Non.Stat.plot <- rbind(as.data.frame(nEGs) %>%
    rownames_to_column(var="Subset") %>%
    pivot_longer(names_to="Omic", values_to="Value", -Subset) %>%
    mutate(Quant="EG counts [N]"),
  as.data.frame(percEGs) %>%
    rownames_to_column(var="Subset") %>%
    pivot_longer(names_to="Omic", values_to="Value", -Subset) %>%
    mutate(Quant="EG weight [%]", Value=Value*100)) %>%
  mutate(Subset=short_names_NonStatES[Subset]) %>%
  mutate(Subset=factor(Subset, levels=rev( c("ORFs",
                           "Tax Phylum", "Tax Class", "Tax Order", "Tax Family", "Tax Genus", "Tax Species",
                           "Fun KO", "Fun Pathway"))),
         Omic=factor(Omic, levels=c("MG", "MT", "MP"))) %>%
  ggplot(aes(x=Subset, y=Value, fill=Subset)) +
  geom_bar(stat="identity") +
  coord_flip() +
  facet_grid(Omic~Quant, scales="free_x") + 
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf),
            colour="black", fill=NA, inherit.aes=FALSE, size=0.5) +
  theme_classic() +
  theme(panel.grid.major = element_line(colour="grey", size=0.25, linetype="dashed"),
        legend.position = "none", panel.grid.major.y = element_blank()) +
  scale_y_continuous(expand=c(0.01, 0.01), breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1))))) +
  scale_fill_viridis(discrete=T, option="viridis")

Non.Stat.plot

ggsave(paste0(wd, "/results/figures/NonStatEGsComparison_corrMRS.png"), Non.Stat.plot, dpi=320, height=5, width=6)
ggsave(paste0(wd, "/results/figures/NonStatEGsComparison_corrMRS.svg"), Non.Stat.plot, dpi=320, height=5, width=6)

```

### Eigenvalue plots

```{r Eigenvalue plots}

omic_type <- c("MG", "MT", "MP")
fun_lvl <- paste("fun", c("C_name", "D_name"), sep=".")
tax_lvl <- paste("tax", c("phylum", "class", "order", "family", "genus", "species"), sep=".")
subsets <- c(fun_lvl, tax_lvl)

EGval.comp <- data.frame(omic_type=c(), subset=c(), index=c(), value=c(), NonStat=c())

for(i in omic_type){
  for(j in c("raw", subsets)){
    if(startsWith(i, "MO")){
      set_cycle <- paste(i, j, "NonStatEG", sep=".")
      EGval_cycle <- paste(i, j, "train.svd", sep=".")
    } else {
      set_cycle <- paste(i, j, "NonStatEG.corr", sep=".")
      EGval_cycle <- paste(i, j, "train.svd.corr", sep=".")
    }
    tmp.mat <- get(EGval_cycle)$d
    N <- length(tmp.mat)
    tmp.NonStat <- get(set_cycle)
    
    tmp.df <- data.frame(omic_type=rep(i, N), subset=rep(j, N), index=1:N, value=tmp.mat, NonStat=(((1:N)%in%tmp.NonStat)+0))
    
    EGval.comp <- rbind(EGval.comp, tmp.df)
  }
}

first.EVs <- EGval.comp %>%
  group_by(omic_type, subset) %>%
  mutate(s=sum(value), value.frac=value/s,
         subset=if_else(subset=="raw", ".raw", subset)) %>%
  filter(index==1) %>%
  pull(value.frac)

round(mean(first.EVs)*100, digits=2)
round(sd(first.EVs)*100, digits=2)

EGval.comp %>%
  group_by(omic_type, subset) %>%
  mutate(s=sum(value), value.frac=value/s,
         subset=if_else(subset=="raw", ".raw", subset)) %>%
  separate(subset, into=c("to_delete", "subset", remove=T, sep=".")) %>%
  mutate(subset=case_when(subset=="C" ~ "Pathway",
                          subset=="D" ~ "Reaction",
                          subset=="raw" ~ "ORFs",
                          .default = subset)) %>%
  ggplot(aes(x=index, y=value.frac, group=1)) +
  geom_point() +
  geom_line() +
  facet_grid(subset~omic_type, scales="free") +
  theme_minimal() +
  labs(x="EG index", y="Fractional eigevalue")

EGval.comp %>%
  filter(index!=1) %>%
  group_by(omic_type, subset) %>%
  mutate(s=sum(value), value.frac=value/s,
         subset=if_else(subset=="raw", ".raw", subset)) %>%
  separate(subset, into=c("to_delete", "subset", remove=T, sep=".")) %>%
  mutate(subset=case_when(subset=="C" ~ "Pathway",
                          subset=="D" ~ "Reaction",
                          subset=="raw" ~ "ORFs",
                          .default = subset)) %>%
  ggplot(aes(x=index, y=value.frac, group=1)) +
  geom_point() +
  geom_line() +
  facet_grid(subset~omic_type, scales="free") +
  theme_minimal() +
  labs(x="EG index", y="Fractional eigevalue")

EGval.comp %>%
  filter(index!=1) %>%
  mutate(NonStat=if_else(NonStat==1, "Time.dependent", "Time.independent")) %>%
  group_by(omic_type, subset) %>%
  mutate(s=sum(value), value.frac=value/s,
         subset=if_else(subset=="raw", ".raw", subset)) %>%
  separate(subset, into=c("to_delete", "subset", remove=T, sep=".")) %>%
  mutate(subset=case_when(subset=="C" ~ "Pathway",
                          subset=="D" ~ "Reaction",
                          subset=="raw" ~ "ORFs",
                          .default = subset)) %>%
  ggplot(aes(x=index, y=value.frac, group=1)) +
  geom_point(aes(color=NonStat)) +
  geom_line() +
  facet_grid(subset~omic_type, scales="free") +
  theme_minimal() +
  labs(x="EG index", y="Fractional eigevalue")

```

## EG analysis

### Time-dependent EG correlations

```{r Time-dependent EG correlations}

omic_type <- c("MG", "MT", "MP")
fun_lvl <- paste("fun", c("C_name", "D_name"), sep=".")
tax_lvl <- paste("tax", c("phylum", "class", "order", "family", "genus", "species"), sep=".")
subsets <- c(fun_lvl, tax_lvl)
set_type <- paste0(c((expand.grid(a=omic_type, b=subsets) %>% mutate(d=paste(a, b, sep=".")))$d, "MG.raw.a", "MT.raw.a", "MP.raw.a"))
set_type_dd <- paste0(c((expand.grid(a=omic_type, b=subsets) %>% mutate(d=paste(a, b, sep=".")))$d, "MG.raw", "MT.raw", "MP.raw"), ".train.svd.corr")

all.EGs <- t(impute_weekly(t(do.call(cbind, lapply(as.list(set_type_dd), function(x) get(x)$v))), colnames(MT.raw.train.clr.corr)))
all.EVs <- unlist(lapply(as.list(set_type_dd), function(x) get(x)$d))
EG_names <- unlist(lapply(set_type_dd, function(x) paste0(paste(unlist(str_split(x, "\\."))[1:3], collapse="."), ".EG", 1:ncol(get(x)$v))))
EG_names <- str_replace(EG_names, "train", "a")

EVs.dd <- data.frame(value=all.EVs, EG_name=EG_names) %>%
  separate(EG_name, into=c("Omic", "Set", "Subset", "EG_num"), sep="\\.", remove=F) %>%
  group_by(Omic, Set, Subset) %>%
  mutate(ES=sum(value), value.perc=value/ES, value.censored=value)
EVs.dd[EVs.dd$EG_num=="EG1","value.censored"] <- 0
EVs.dd <- EVs.dd %>%
  group_by(Omic, Set, Subset) %>%
  mutate(ES.scaled=sum(value.censored), value.scaled=value.censored/ES.scaled)

subsets <- c(fun_lvl, tax_lvl)
set_raw <- c("raw")
set_old_type <- paste0(c((expand.grid(a=omic_type, b=subsets) %>% mutate(d=paste(a, b, sep=".")))$d, "MG.raw.a", "MT.raw.a", "MP.raw.a"))
set_type <- c(paste0(c((expand.grid(a=omic_type, b=subsets) %>% mutate(d=paste(a, b, sep=".")))$d)))
set_raw_type <- paste0(c((expand.grid(a=omic_type, b=set_raw) %>% mutate(d=paste(a, b, sep=".")))$d))

non_stat_tot <- c(unlist(sapply(set_type, function(x) paste(x, ".EG", get(paste(x, "NonStatEG.corr", sep=".")), sep=""))),
                  unlist(sapply(set_raw_type, function(x) paste(x, ".EG", get(paste(x, "NonStatEG.corr", sep=".")), sep=""))))

non_stat_tot_a <- c(unlist(sapply(set_type, function(x) paste(x, ".EG", get(paste(x, "NonStatEG.corr", sep=".")), sep=""))),
                  unlist(sapply(set_raw_type, function(x) paste(x, ".a.EG", get(paste(x, "NonStatEG.corr", sep=".")), sep=""))))

EG_names <- unlist(lapply(set_type_dd, function(x) paste0(paste(unlist(str_split(x, "\\."))[1:3], collapse="."), ".EG", 1:ncol(get(x)$v))))
EG_names <- str_replace(EG_names, "train", "a")

Adj0 <- cor((all.EGs[1:nrow(all.EGs),] %>%
               `colnames<-`(EG_names))[,non_stat_tot_a],
            (all.EGs[1:nrow(all.EGs),] %>%
               `colnames<-`(EG_names))[,non_stat_tot_a],
            method="pearson")

```


### EG dist comparison

```{r, EGs dist comparison}

EG.dist.euclidean <- dist((Adj0), method="euclidean")
EG.dist.maximum <- dist((Adj0), method="maximum")
EG.dist.minkowski <- dist((Adj0), method="minkowski")
EG.dist.manhattan <- dist((Adj0), method="manhattan")
EG.dist.canberra <- dist((Adj0), method="canberra")

EG.dist.list <- list(EG.dist.euclidean, EG.dist.maximum, EG.dist.minkowski, EG.dist.manhattan, EG.dist.canberra)
N <- length(EG.dist.list)
EG.dist.mat <- matrix(nrow=N, ncol=N) %>%
  `colnames<-`(c("euclidean", "maximum", "minkowski", "manhattan", "canberra")) %>%
  `rownames<-`(c("euclidean", "maximum", "minkowski", "manhattan", "canberra"))
EG.dist.ps <- matrix(nrow=N, ncol=N) %>%
  `colnames<-`(c("euclidean", "maximum", "minkowski", "manhattan", "canberra")) %>%
  `rownames<-`(c("euclidean", "maximum", "minkowski", "manhattan", "canberra"))

for(i in 1:N){
  for(j in 1:N){
    EG.dist.mat[i,j] <- mantel.rtest(EG.dist.list[[i]], EG.dist.list[[j]])$obs
    EG.dist.ps[i,j] <- mantel.rtest(EG.dist.list[[i]], EG.dist.list[[j]])$pvalue
  }
}

corrplot::corrplot.mixed(EG.dist.mat)
EG.dist.ps %>%
  as.data.frame() %>%
  knitr::kable()

```

### EG clustering

```{r, EGs clustering}

EG.dist <- as.matrix(dist((Adj0), method="minkowski")) #euclidean

EG.hc <- hclust(as.dist(EG.dist), method="complete");
sizeGrWindow(12,9)
plot(EG.hc, xlab="", sub="", main="Gene clustering on TOM-based dissimilarity", labels=F, hang=0.04)

minModuleSize <- 3
dynamicMods = cutreeDynamic(dendro=EG.hc, distM=EG.dist, deepSplit=0, pamRespectsDendro=FALSE, minClusterSize=minModuleSize)
table(dynamicMods)

dynamicColors <- labels2colors(dynamicMods)
table(dynamicColors)

sizeGrWindow(8,6)
plotDendroAndColors(EG.hc, dynamicColors, "Dynamic Tree Cut", dendroLabels=F, hang=0.03, addGuide=T, guideHang=0.05, main="Eigengene clustering")

```

### Further EG clustering

```{r EG clustering 2}

MEList = moduleEigengenes(abs(Adj0), colors=dynamicColors)
MEs = MEList$eigengenes
MEDiss = 1-cor(MEs);
METree = hclust(as.dist(MEDiss), method="complete");
sizeGrWindow(7, 6)
plot(METree, main = "Clustering of module eigengenes", xlab = "", sub = "")

MEDissThres = 0.0
abline(h=MEDissThres, col = "red")
merge = mergeCloseModules(abs(Adj0), dynamicColors, cutHeight=MEDissThres, verbose = 3)
mergedColors = merge$colors
mergedMEs = merge$newMEs

```


```{r, merged clusters}

sizeGrWindow(12, 9)
plotDendroAndColors(EG.hc, cbind(dynamicColors, mergedColors), c("Dynamic Tree Cut", "Merged dynamic"), dendroLabels = FALSE, hang = 0.03, addGuide = TRUE, guideHang = 0.05)

colors2num <- factor(paste0("S",1:length(unique(mergedColors))), levels=paste0("S",1:length(unique(mergedColors))))
names(colors2num) <- unique(mergedColors)

write_tsv(data.frame(EG_name=non_stat_tot_a, module=mergedColors, mod_num=colors2num[mergedColors]),
          file=paste0(wd, "/results/tables/EG_clustringMRS"))

```

### Upset plot

```{r upset plot}

library(ggupset)
library(ggbreak)

to_upset <- (data.frame(Origin=non_stat_tot_a, module=dynamicColors, EV=EVs.dd[EVs.dd$EG_name%in%non_stat_tot_a, "value.scaled"])) %>%
  separate(Origin, into=c("omic", "category", "subcategory", "EGnum"), sep="\\.", remove=F) %>%
  group_by(omic, category, subcategory) %>%
  mutate(value.sum=sum(value.scaled), EV=value.scaled/value.sum) %>%
  mutate(EV=value.scaled)

EG_order <- (to_upset %>%
  mutate(EG_type=paste(omic, subcategory, sep=" ")) %>%
  group_by(module) %>%
  mutate(EG_type = list(EG_type)) %>%
  mutate(EV_sum = max(EV)) %>%
  ungroup() %>%
  arrange(EV_sum) %>%
  distinct(EG_type, EV_sum) %>%
  dplyr::select(EV_sum, EG_type))$EV_sum

sum(EG_order)

upset.plot <- to_upset %>%
  mutate(subcategory=case_when(subcategory=="a" ~ "ORF",
                               subcategory=="C_name" ~ "pathway",
                               subcategory=="D_name" ~ "reaction",
                               T ~ subcategory)) %>%
  mutate(EG_type=paste(omic, subcategory, sep=" ")) %>%
  group_by(module) %>%
  mutate(EG_type=list(EG_type)) %>%
  ungroup() %>%
  ggplot(aes(x=EG_type, y=EV)) +
  geom_violin() +
  stat_summary(fun.y="max", geom="point", color="red") +
  scale_y_continuous(breaks=c(0, 2, 5, 10, 30, 60)) +
  scale_x_upset(n_intersections=Inf, ytrans="sqrt") +
  theme_classic()

upset.plot

ggsave(paste0(wd, "/results/figures/upsetMRS.png"), upset.plot, dpi=320, height=6, width=7)
ggsave(paste0(wd, "/results/figures/upsetMRS.svg"), upset.plot, dpi=320, height=6, width=7)

```

### Non Stat EG comp

```{r Non Stat EG comp}

short_names_NonStatES <- c("ORFs",
                           "Tax Phylum", "Tax Class", "Tax Order", "Tax Family", "Tax Genus", "Tax Species",
                           "Fun KO", "Fun Pathway")
names(short_names_NonStatES) <- c("raw.a",
                                        "tax.phylum", "tax.class", "tax.order", "tax.family", "tax.genus", "tax.species",
                                        "fun.D_name", "fun.C_name")

to_upset %>%
  group_by(omic, category, subcategory) %>%
  mutate(count_EG=n(), EV_sum=sum(value.scaled)) %>%
  dplyr::select(-c(EV_sum, EV)) %>%
  ungroup() %>%
  dplyr::select(omic, category, subcategory, value.scaled) %>%
  pivot_longer(names_to="measure", values_to="value", -c(omic, category, subcategory)) %>%
  mutate(Subset=paste0(category, ".", subcategory), Subset=short_names_NonStatES[Subset]) %>%
  mutate(Subset=factor(Subset, levels=rev( c("ORFs",
                           "Tax Phylum", "Tax Class", "Tax Order", "Tax Family", "Tax Genus", "Tax Species",
                           "Fun KO", "Fun Pathway"))),
         omic=factor(omic, levels=c("MG", "MT", "MP"))) %>%
  ggplot(aes(x=Subset, y=value)) +
  geom_bar(stat="identity") +
  coord_flip() +
  facet_grid(omic~measure, scales="free_x") + 
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf),
            colour="black", fill=NA, inherit.aes=FALSE, size=0.5) +
  theme_classic() +
  theme(panel.grid.major = element_line(colour="grey", size=0.25, linetype="dashed"),
        legend.position = "none", panel.grid.major.y = element_blank()) +
  scale_y_continuous(expand=c(0.01, 0.01), breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1))))) +
  scale_fill_viridis(discrete=T, option="viridis")

```


### rEG selection

```{r rEG selection}

noise.dd <- data.frame(EG_name=non_stat_tot_a, module=mergedColors, noise=colSds(abs((all.EGs[2:nrow(all.EGs),] %>% `colnames<-`(EG_names))[,non_stat_tot_a] - (all.EGs[1:(nrow(all.EGs)-1),] %>% `colnames<-`(EG_names))[,non_stat_tot_a])), EV=to_upset$EV)

noise.ff <- noise.dd %>%
  filter(!startsWith(EG_name, "MP")) %>%
  mutate(weighted_noise=noise*(1-EV)) %>%
  separate(EG_name, into=c("omic", "category", "subcategory", "EG_num"), sep="\\.", remove=F) %>%
  separate(EG_num, into=c("to_delete", "EG_num"), sep="EG", remove=T) %>%
  dplyr::select(-to_delete) %>%
  filter(EG_num>1) %>%
  group_by(module) %>%
  summarise(min_noise=min(noise), rep_EG=EG_name[which(noise==min_noise)])

noise.ff

```

### rEG Explained Variance

```{r rEG Explained Variance}

signals.EVs <- to_upset %>%
  ungroup() %>%
  mutate(module=colors2num[module]) %>%
  dplyr::select(module, EV) %>%
  group_by(module) %>%
  summarise(EV=max(EV))

signals.EVs

sum(signals.EVs$EV)

write.csv(file=paste0(wd, "/results/tables/signals_EVs.csv"), as.data.frame(signals.EVs), row.names=F)

```


### rEG plot

```{r rEG plot}

Sys.setenv("LANGUAGE"="En")
Sys.setlocale("LC_ALL", "en_US.UTF-8")

sampled_weeks <- as.numeric(floor(difftime(ymd(colnames(MG.raw.train.clr.corr)), ymd(colnames(MG.raw.train.clr.corr))[1], units="week"))+1)

non_stat2mod <- mergedColors
names(non_stat2mod) <- non_stat_tot_a

signals.plot <- as.data.frame((all.EGs[sampled_weeks, ] %>% `colnames<-`(EG_names))[,non_stat_tot_a]) %>%
  `rownames<-`(colnames(MG.raw.train.clr.corr)) %>%
  rownames_to_column(var="Sample") %>%
  pivot_longer(names_to="EG_name", values_to="value", -c(Sample)) %>%
  mutate(module=non_stat2mod[EG_name]) %>%
  separate(EG_name, into=c("omic", "category", "subcategory", "EG_num"), sep="\\.", remove=F) %>%
  mutate(module=colors2num[non_stat2mod[EG_name]]) %>%
  ggplot(aes(x=as.Date(Sample), y=value, group=EG_name)) +
  geom_point() +
  geom_line() +
  geom_smooth(se=0, alpha=.5) +
  facet_wrap(module~., ncol=4, scales="free") +
  theme_classic()

signals.plot

ggsave(paste0(wd, "/results/figures/signals_traceMRS.png"), signals.plot, dpi=320, height=6, width=10)
ggsave(paste0(wd, "/results/figures/signals_traceMRS.svg"), signals.plot, dpi=320, height=6, width=10)

signals.plot <- as.data.frame((all.EGs[sampled_weeks, ] %>%`colnames<-`(EG_names))[,noise.ff$rep_EG]) %>%
  `rownames<-`(colnames(MG.raw.train.clr.corr)) %>%
  rownames_to_column(var="Sample") %>%
  pivot_longer(names_to="EG_name", values_to="value", -Sample) %>%
  separate(EG_name, into=c("omic", "category", "subcategory", "EG_num"), sep="\\.", remove=F) %>%
  mutate(module=colors2num[non_stat2mod[EG_name]]) %>%
  ggplot(aes(x=as.Date(Sample), y=value, group=EG_name)) +
  geom_point() +
  geom_line() +
  facet_wrap(module~., ncol=4, scales="free_y") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

as.data.frame((all.EGs[sampled_weeks, ] %>% `colnames<-`(EG_names))[,noise.ff$rep_EG]) %>%
  `rownames<-`(colnames(MG.raw.train.clr.corr)) %>%
  rownames_to_column(var="Sample") %>%
  pivot_longer(names_to="EG_name", values_to="value", -Sample) %>%
  separate(EG_name, into=c("omic", "category", "subcategory", "EG_num"), sep="\\.", remove=F) %>%
  mutate(module=colors2num[non_stat2mod[EG_name]]) %>%
  ggplot(aes(x=as.Date(Sample), y=value, group=EG_name)) +
  geom_point() +
  geom_line() +
  facet_wrap(module~., ncol=4, scales="free_y") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

signals.plot

ggsave(paste0(wd, "/results/figures/signalsMRS.png"), signals.plot, dpi=320, height=4.5, width=10)
ggsave(paste0(wd, "/results/figures/signalsMRS.svg"), signals.plot, dpi=320, height=4.5, width=10)

```


### Omic composition

```{r, Omic composition}

as.data.frame(all.EGs[sampled_weeks, ] %>%
                `colnames<-`(EG_names)) %>%
  `rownames<-`(colnames(MG.raw.train.clr.corr)) %>%
  rownames_to_column(var="Sample") %>%
  pivot_longer(names_to="EG_name", values_to="value", -Sample) %>%
  separate(EG_name, into=c("omic", "category", "subcategory", "EG_num"), sep="\\.", remove=F) %>%
  mutate(module=colors2num[non_stat2mod[EG_name]]) %>%
  filter(!is.na(module)) %>%
  dplyr::select(c("EG_name", "omic", "module")) %>%
  distinct() %>%
  ungroup() %>%
  group_by(omic, module) %>%
  summarise(n=n()) %>%
  ungroup() %>%
  pivot_wider(names_from=module, values_from=n) %>%
  column_to_rownames(var="omic")

```


### Granger causality

```{r, Granger causality}

granger.mat <- matrix(ncol=length(noise.ff$rep_EG), nrow=length(noise.ff$rep_EG)) %>%
  `colnames<-`(noise.ff$rep_EG) %>%
  `rownames<-`(noise.ff$rep_EG)

weeks.EGs <- as.data.frame((all.EGs[sampled_weeks, ] %>% `colnames<-`(EG_names))[,noise.ff$rep_EG])

weeks.EGs[!1:nrow(all.EGs)%in%sampled_weeks, ] <- NA
weekly_imputed <- t(impute_weekly(t(weeks.EGs), colnames(MT.raw.train.clr.corr)))
  
for(i in rownames(granger.mat)){
  for(j in colnames(granger.mat)){
    if(i != j){
      granger.mat[i,j] <- (grangertest(weekly_imputed[,noise.ff$rep_EG][,i],
                                       weekly_imputed[,noise.ff$rep_EG][,j],
                                       order=10, na.action=na.omit))$`Pr(>F)`[2]
    }
  }
}

sum(granger.mat<0.05, na.rm=T)

```

### Nonlinearity checks

```{r, Nonlinearity checks}

EG.mat.wi <- weekly_imputed[,noise.ff$rep_EG]
colnames(EG.mat.wi) <- colors2num[non_stat2mod[colnames(EG.mat.wi)]]

vars <- colnames(EG.mat.wi)
simplex_out <- lapply(vars, function(var) {
    simplex(EG.mat.wi[,var], E=1:20)
})
names(simplex_out) <- vars

embedding.plot <- bind_rows(simplex_out) %>%
  mutate(signal=unlist(lapply(names(simplex_out), rep, 20)),
         signal=factor(signal, levels=paste0("S", 1:17)),
         rho=unlist(rho)) %>%
  ggplot(aes(x=E, y=rho)) +
  geom_line() +
  facet_wrap(signal~., ncol=6, scales="free") +
  theme_classic() +
  labs(x="Embedding dimensions (E)", y="Forecast Skill (rho)")

embedding.plot
ggsave(paste0(wd, "/results/figures/embedding.png"), embedding.plot, dpi=320, height=4.5, width=10)
ggsave(paste0(wd, "/results/figures/embedding.svg"), embedding.plot, dpi=320, height=4.5, width=10)

best_E <- sapply(simplex_out, function(df) {
    df$E[which.max(df$rho)]
})
best_E

smap_out <- lapply(vars, function(var) {
    s_map(EG.mat.wi[,var], E = best_E[var])
})
names(smap_out) <- names(simplex_out)

nonlinearity.plot <- bind_rows(smap_out) %>%
  mutate(signal=unlist(lapply(names(simplex_out), rep, 18)),
         signal=factor(signal, levels=paste0("S", 1:17)),
         rho=unlist(rho)) %>%
  ggplot(aes(x=theta, y=rho)) +
  geom_line() +
  facet_wrap(signal~., ncol=6, scales="free") +
  theme_classic() +
  labs(x="Nonlinearity (theta)", y="Forecast Skill (rho)")

nonlinearity.plot
ggsave(paste0(wd, "/results/figures/nonlinearity.png"), nonlinearity.plot, dpi=320, height=4.5, width=10)
ggsave(paste0(wd, "/results/figures/nonlinearity.svg"), nonlinearity.plot, dpi=320, height=4.5, width=10)

```

### Convergent Cross Mapping validation of nonlinear links

```{r, Convergent Cross Mapping validation of nonlinear links}

S7_xmap_S8 <- ccm(EG.mat.wi, lib = segments, pred = segments, lib_column = "S7", 
    target_column = "S8", E = best_E["S7"], silent = T, replace = T, stats_only = T, lib_sizes=c(20, 50, 1))
S4_xmap_S8 <- ccm(EG.mat.wi, lib = segments, pred = segments, lib_column = "S4", 
    target_column = "S8", E = best_E["S4"], silent = T, replace = T, stats_only = T, lib_sizes=c(20, 50, 1))
S9_xmap_S8 <- ccm(EG.mat.wi, lib = segments, pred = segments, lib_column = "S9", 
    target_column = "S8", E = best_E["S9"], silent = T, replace = T, stats_only = T, lib_sizes=c(20, 50, 1))
S10_xmap_S17 <- ccm(EG.mat.wi, lib = segments, pred = segments, lib_column = "S10", 
    target_column = "S17", E = best_E["S10"], silent = T, replace = T, stats_only = T, lib_sizes=c(20, 50, 1))
S1_xmap_S17 <- ccm(EG.mat.wi, lib = segments, pred = segments, lib_column = "S1", 
    target_column = "S17", E = best_E["S1"], silent = T, replace = T, stats_only = T, lib_sizes=c(20, 50, 1))
S6_xmap_S8 <- ccm(EG.mat.wi, lib = segments, pred = segments, lib_column = "S6", 
    target_column = "S7", E = best_E["S6"], silent = T, replace = T, stats_only = T, lib_sizes=c(20, 50, 1))
S5_xmap_S8 <- ccm(EG.mat.wi, lib = segments, pred = segments, lib_column = "S5", 
    target_column = "S7", E = best_E["S5"], silent = T, replace = T, stats_only = T, lib_sizes=c(20, 50, 1))

CCM.plot <- rbind((S7_xmap_S8 %>%
  pivot_longer(names_to="Direction", -c(LibSize, E, tau, tp, nn))),
  (S4_xmap_S8 %>%
  pivot_longer(names_to="Direction", -c(LibSize, E, tau, tp, nn))),
  (S9_xmap_S8 %>%
  pivot_longer(names_to="Direction", -c(LibSize, E, tau, tp, nn))),
  (S10_xmap_S17 %>%
  pivot_longer(names_to="Direction", -c(LibSize, E, tau, tp, nn))),
  (S1_xmap_S17 %>%
  pivot_longer(names_to="Direction", -c(LibSize, E, tau, tp, nn))),
  (S6_xmap_S8 %>%
  pivot_longer(names_to="Direction", -c(LibSize, E, tau, tp, nn))),
  (S5_xmap_S8 %>%
  pivot_longer(names_to="Direction", -c(LibSize, E, tau, tp, nn)))) %>%
  dplyr::select(LibSize, Direction, value) %>%
  mutate(Edge=lapply(str_split(Direction, ":"), sort)) %>%
  mutate(Direction=gsub(pattern=":", replacement="->", x=Direction)) %>%
  rowwise() %>%
  mutate(Edge=paste(Edge, collapse="-")) %>%
  ggplot(aes(x=LibSize, y=value, group=Direction, color=Direction)) +
  geom_line() +
  theme_classic() +
  facet_wrap(Edge~., nrow=3, scales="free") +
  labs(x="Library Size", y="Cross Map Skill (rho)")

CCM.plot
ggsave(paste0(wd, "/results/figures/CCM.png"), CCM.plot, dpi=320, height=4.5, width=7)
ggsave(paste0(wd, "/results/figures/CCM.svg"), CCM.plot, dpi=320, height=4.5, width=7)
```

### Granger causality persisten homology

```{r, Granger causality persisten homology}

weeks.EGs <- as.data.frame((all.EGs[sampled_weeks, ] %>% `colnames<-`(EG_names))[,noise.ff$rep_EG])
  
weeks.EGs[!1:nrow(all.EGs)%in%sampled_weeks, ] <- NA
weekly_imputed <- t(impute_weekly(t(weeks.EGs), colnames(MT.raw.train.clr.corr)))

for(i in rownames(granger.mat)){
  for(j in colnames(granger.mat)){
    if(i != j){
      granger.mat[i,j] <- (grangertest(weekly_imputed[,noise.ff$rep_EG][,i],
                                       weekly_imputed[,noise.ff$rep_EG][,j],
                                       order=1, na.action=na.omit))$`Pr(>F)`[2]
    }
  }
}

granger.aggregate <- as.data.frame(granger.mat) %>%
    rownames_to_column(var="Origin") %>%
    pivot_longer(names_to="Target", values_to="lag_1", -c(Origin)) %>%
    mutate(Origin=colors2num[non_stat2mod[Origin]], Target=colors2num[non_stat2mod[Target]])

for(lag.cycle in 2:16){
  granger.mat <- matrix(ncol=length(noise.ff$rep_EG), nrow=length(noise.ff$rep_EG)) %>%
    `colnames<-`(noise.ff$rep_EG) %>%
    `rownames<-`(noise.ff$rep_EG)
    
  for(i in rownames(granger.mat)){
    for(j in colnames(granger.mat)){
      if(i != j){
        granger.mat[i,j] <- (grangertest(weekly_imputed[,noise.ff$rep_EG][,i],
                                         weekly_imputed[,noise.ff$rep_EG][,j],
                                         order=lag.cycle, na.action=na.omit))$`Pr(>F)`[2]
      }
    }
  }
  
  granger.long <- as.data.frame(granger.mat) %>%
    rownames_to_column(var="Origin") %>%
    pivot_longer(names_to="Target", values_to=paste0("lag_", lag.cycle), -c(Origin)) %>%
    mutate(Origin=colors2num[non_stat2mod[Origin]], Target=colors2num[non_stat2mod[Target]])
  
  granger.aggregate <- left_join(granger.aggregate, granger.long, by=c("Origin", "Target"))
  
}

granger.aggregate %>%
  filter(!is.na(lag_1)) %>%
  pivot_longer(names_to="lags", values_to="value", -c(Origin, Target)) %>%
  filter(value<0.05) %>%
  ggplot(aes(y=factor(lags, levels=paste0("lag_", 1:16)), x=paste0(Origin, " ", Target), fill=value)) +
  geom_tile() +
  theme_classic()

to_cytoscape <- granger.aggregate %>%
  filter(!is.na(lag_1)) %>%
  pivot_longer(names_to="lags", values_to="value", -c(Origin, Target)) %>%
  mutate(value=ifelse(value<0.05, 1, 0), OriTar=paste0(Origin, Target), lags=factor(lags, levels=paste0("lag_", 1:16))) %>%
  group_by(OriTar, grp=with(rle(value), rep(seq_along(lengths), lengths))) %>%
  mutate(Counter=seq_along(grp)) %>%
  ungroup() %>%
  dplyr::select(-grp) %>%
  mutate(Counter=Counter*value) %>%
  group_by(OriTar) %>%
  mutate(max_count=max(Counter)) %>%
  ungroup() %>%
  dplyr::select(-c(value, OriTar, lags, Counter)) %>%
  filter(max_count>=6) %>%
  distinct()

dim(to_cytoscape)

write_csv(to_cytoscape, paste0(wd, "/results/tables/Signal_netPH.csv"))

```

### Granger net

```{r, Granger net}

to_cytoscape <- as.data.frame(granger.mat) %>%
  rownames_to_column(var="Origin") %>%
  pivot_longer(names_to="Target", values_to="value", -c(Origin)) %>%
  mutate(Origin=colors2num[non_stat2mod[Origin]], Target=colors2num[non_stat2mod[Target]]) %>%
  filter(value<0.05)

hist(to_cytoscape$value)# better with filter value>0

write_csv(to_cytoscape, paste0(wd, "/results/tables/Signal_netMRS_Min.csv"))

```


### Saving image

```{r, Saving image}

save.image(file=paste0(wd, "/preprocessed/data_stable_MRS.RData"))

```


## EG enrichment

### EG enrichment

```{r EG enrichment}

dd_quantile <- data.frame(omic="", mat_name="", EG_num="", direction="", name="")

for(i in non_stat_tot_a){
  splitted_cycle <- unlist(str_split(i, "\\."))
  EG_num_cycle <- as.numeric(unlist(str_split(splitted_cycle[4], "EG"))[2])
  if(splitted_cycle[3]=="a"){
    names_cycle <- rownames(get(paste0(splitted_cycle[1], ".", splitted_cycle[2], ".train.clr.corr")))
    arr_cycle <- get(paste0(splitted_cycle[1], ".", splitted_cycle[2], ".train.svd.corr"))$u[,EG_num_cycle]
  } else {
    names_cycle <- rownames(get(paste0(splitted_cycle[1], ".", splitted_cycle[2], ".", splitted_cycle[3], ".train.clr.corr")))
    arr_cycle <- get(paste0(splitted_cycle[1], ".", splitted_cycle[2], ".", splitted_cycle[3], ".train.svd.corr"))$u[,EG_num_cycle]
  }
  names(arr_cycle) <- names_cycle
  
  if(splitted_cycle[3]=="a"){ 
    print("skip")
  } else {
    
    if(splitted_cycle[2]=="tax"){
      count_dd <- LAO_tax %>%
        group_by_at(c(splitted_cycle[3], "ORF")) %>% summarise() %>%
        group_by_at(c(splitted_cycle[3])) %>% summarise(n=n())
      tab_abundance <- unlist(count_dd[,2])
      names(tab_abundance) <- unlist(count_dd[,1])
      arr_cycle <- arr_cycle[names(tab_abundance[tab_abundance>500])]
    }
    
    if(splitted_cycle[2]=="fun"){
      count_dd <- LAO_ko %>%
        group_by_at(c(splitted_cycle[3], "ORF")) %>% summarise() %>%
        group_by_at(c(splitted_cycle[3])) %>% summarise(n=n())
      tab_abundance <- unlist(count_dd[,2])
      names(tab_abundance) <- unlist(count_dd[,1])
      arr_cycle <- arr_cycle[names(tab_abundance[tab_abundance>20])]
    }
    
    up_cycle <- sort(arr_cycle[arr_cycle>quantile(arr_cycle, c(0.95), na.rm=T)], decreasing=T)
    down_cycle <- sort(arr_cycle[arr_cycle<quantile(arr_cycle, c(0.05), na.rm=T)], decreasing=F)
    
    up_n <- length(up_cycle)
    down_n <- length(down_cycle)
    
    up_mat <- data.frame(omic=rep(splitted_cycle[1], up_n), mat_name=rep(paste0(splitted_cycle[2], ".", splitted_cycle[3]), up_n), EG_num=rep(splitted_cycle[4], up_n), direction=rep("up", up_n), name=names(up_cycle))
    down_mat <- data.frame(omic=rep(splitted_cycle[1], down_n), mat_name=rep(paste0(splitted_cycle[2], ".", splitted_cycle[3]), down_n), EG_num=rep(splitted_cycle[4], down_n), direction=rep("down", down_n), name=names(down_cycle))
    
    dd_quantile <- rbind(dd_quantile, rbind(up_mat, down_mat))
    
  }
}

dd_quantile <- dd_quantile[2:nrow(dd_quantile),]

dd_extra <- left_join((dd_quantile %>% `colnames<-`(c("omic", "mat_name", "EGnum", "direction", "D_name"))),
                       (map_ko %>% dplyr::select(D_name, C_name, B_name)))

write_tsv(dd_quantile, paste0(wd, "/results/tables/EGs_percentilesMRS.tsv"))

```
