---
title: "MicroForecast - Script 0"
author: "Francesco Delogu"
date: as.Date(now())
output: html_document
---

# MicroForecast - Script 0

## Set the environment

```{r setup}

options(stringsAsFactors = F, gsubfn.engine = "R")

library(tidyverse)
library(matrixStats)
library(ggridges)
library(compositions)
library(MASS)
library(caret)
library(lubridate)
library(imputeTS)
library(parallel)
library(tseries)
library(forecast)
library(olsrr)
library(fable)
library(viridis)
library(TidyMultiqc)

wd <- getwd()

```


## Load data

### Load counts

```{r load counts}

MG.raw <- (read_tsv(paste0(wd, "/data/MG_harmonized.tsv")) %>%
  column_to_rownames(var="ORF"))

MT.raw <- (read_tsv(paste0(wd, "/data/MT_harmonized.tsv")) %>%
  column_to_rownames(var="ORF"))

MP.raw <- (read_tsv(paste0(wd, "/data/summary_prophane.tsv")) %>%
  mutate(ORF=gsub("Good", "", gsub(";", "+", members_identifier))) %>%
  dplyr::select(-members_identifier) %>%
  column_to_rownames(var="ORF"))

```

### Load metadata

```{r load metadata}

LAO_meta <- read_tsv(paste0(wd, "/data/LAO_metadata.tsv"))
colnames(LAO_meta) <- c("Date", colnames(LAO_meta)[2:length(colnames(LAO_meta))])
LAO_meta$Date <- as.Date(LAO_meta$Date, "%m/%d/%y")
LAO_meta$Oxygen_manual <- as.numeric(LAO_meta$Oxygen_manual)
LAO_meta$Oxygen_sat_manual <- as.numeric(LAO_meta$Oxygen_sat_manual)
LAO_meta$Time[is.na(LAO_meta$Time)] <- median(LAO_meta$Time, na.rm=T)
LAO_meta <- LAO_meta %>%
  mutate(DateTime = ymd_hms(paste(LAO_meta$Date, LAO_meta$Time, sep=" ")))

Sivec_meta <- read_tsv(paste0(wd, "/data/Sivec_Schifflange.csv"))
Sivec_meta$Date <- as.Date(Sivec_meta$Date, "%m/%d/%y")
Sivec_meta <- Sivec_meta %>%
  mutate(DateTime = ymd_h(paste(Sivec_meta$Date, Sivec_meta$Time, sep=" ")),
         Time = paste0(Time, ":00:00"))

LAO_samples <- read.table(paste0(wd, "/data/samples.txt"), sep=" ", header=T)
sample_to_date <- as.Date(LAO_samples$Date)
names(sample_to_date) <- LAO_samples$Sample

MG.multiqc <- load_multiqc(paste0(wd, "/data/MG_multiqc_data.json")) %>%
  separate(metadata.sample_id, into=c("to_delete1", "to_delete2", "to_delete3",
                                      "to_delete4", "to_delete5","to_delete6",
                                      "metadata.sample_id", "to_delete8", "to_delete9",
                                      "to_delete10", "Sample"), sep=" \\| ", remove=T, extra="drop") %>%
  dplyr::select(-c("to_delete1", "to_delete2", "to_delete3",
                   "to_delete4", "to_delete5","to_delete6",
                   "to_delete8", "to_delete9","to_delete10")) %>%
  mutate(metadata.sample_id = sample_to_date[metadata.sample_id]) %>%
  separate(Sample, into=c("to_delete1", "Read_set", "File_type"), sep="\\.", remove=T, extra="drop") %>%
  mutate(Read_set=toupper(Read_set)) %>%
  dplyr::select(-c("to_delete1")) %>%
  arrange(metadata.sample_id)

MT.multiqc <- load_multiqc(paste0(wd, "/data/MT_multiqc_data.json")) %>%
  separate(metadata.sample_id, into=c("to_delete1", "to_delete2", "to_delete3",
                                      "to_delete4", "to_delete5","to_delete6",
                                      "metadata.sample_id", "to_delete8", "to_delete9",
                                      "to_delete10", "Sample"), sep=" \\| ", remove=T, extra="drop") %>%
  dplyr::select(-c("to_delete1", "to_delete2", "to_delete3",
                   "to_delete4", "to_delete5","to_delete6",
                   "to_delete8", "to_delete9","to_delete10")) %>%
  mutate(metadata.sample_id = sample_to_date[metadata.sample_id]) %>%
  separate(Sample, into=c("to_delete1", "Read_set", "File_type"), sep="\\.", remove=T, extra="drop") %>%
  mutate(Read_set=toupper(Read_set)) %>%
  dplyr::select(-c("to_delete1")) %>%
  arrange(metadata.sample_id)

```

### Load functional annotation

```{r load functional annotation}

LAO_ko <- read_tsv(paste0(wd, "/data/mantis_integrated.kofam"), col_names=c("ORF", "D_code"))

map_ko <- read_tsv(paste0(wd, "/data/ko.txt"))
LAO_ko <- inner_join(LAO_ko, map_ko)

```

### Load taxonomical annotation

```{r load taxonomical annotation}

gene.map <- read_tsv(paste0(wd, "/data/gene.map5plus"), col_names=F) %>%
  `colnames<-`(c("ORF", "contig", "subset"))

bin.tax <- read_tsv(paste0(wd, "/data/out.BAT.classification")) %>%
  `colnames<-`(c("bin", colnames(.)[-1])) %>%
  separate(superkingdom, c("superkingdom", "to_delete"), sep=":") %>%
  separate(phylum, c("phylum", "to_delete"), sep=":") %>%
  separate(class, c("class", "to_delete"), sep=":") %>%
  separate(order, c("order", "to_delete"), sep=":") %>%
  separate(family, c("family", "to_delete"), sep=":") %>%
  separate(genus, c("genus", "to_delete"), sep=":") %>%
  separate(species, c("species", "to_delete"), sep=":") %>%
  dplyr::select(-to_delete, -lineage, -`lineage scores`, -reason, -classification)

bin.map <- read_tsv(paste0(wd, "/data/bin_summary.txt"), col_names=F) %>%
  `colnames<-`(c("contig", "bin"))

contig.tax <- read_tsv(paste0(wd, "/data/out.CAT2.classification")) %>%
  `colnames<-`(c("contig", colnames(.)[-1])) %>%
  separate(superkingdom, c("superkingdom", "to_delete"), sep=":") %>%
  separate(phylum, c("phylum", "to_delete"), sep=":") %>%
  separate(class, c("class", "to_delete"), sep=":") %>%
  separate(order, c("order", "to_delete"), sep=":") %>%
  separate(family, c("family", "to_delete"), sep=":") %>%
  separate(genus, c("genus", "to_delete"), sep=":") %>%
  separate(species, c("species", "to_delete"), sep=":") %>%
  dplyr::select(-to_delete, -lineage, -`lineage scores`, -reason)

contig.bin.tax <- full_join(bin.map, bin.tax)
rm(bin.tax, bin.map)

euk.bin.tax <- read_tsv(paste0(wd, "/data/euk.BAT.classification")) %>%
  `colnames<-`(c("bin", colnames(.)[-1])) %>%
  separate(superkingdom, c("superkingdom", "to_delete"), sep=":") %>%
  separate(phylum, c("phylum", "to_delete"), sep=":") %>%
  separate(class, c("class", "to_delete"), sep=":") %>%
  separate(order, c("order", "to_delete"), sep=":") %>%
  separate(family, c("family", "to_delete"), sep=":") %>%
  separate(genus, c("genus", "to_delete"), sep=":") %>%
  separate(species, c("species", "to_delete"), sep=":") %>%
  dplyr::select(-to_delete, -lineage, -`lineage scores`, -reason, -classification)

euk.bin.map <- read_tsv(paste0(wd, "/data/euk_summary.txt"), col_names=F) %>%
  `colnames<-`(c("contig", "bin"))

euk.tax <- read_tsv(paste0(wd, "/data/euk_tax.tsv"), col_names=F) %>%
  `colnames<-`(c("contig", "to_delete", "to_delete", "to_delete", "to_delete", "to_delete", "to_delete", "to_delete", "tax")) %>%
  dplyr::select(-to_delete) %>%
  separate(tax, c("superkingdom", "phylum", "class", "order", "family", "genus", "species"), sep=";") %>%
  separate(superkingdom, c("to_delete", "superkingdom"), sep="_") %>%
  separate(phylum, c("to_delete", "phylum"), sep="_") %>%
  separate(class, c("to_delete", "class"), sep="_") %>%
  separate(order, c("to_delete", "order"), sep="_") %>%
  separate(family, c("to_delete", "family"), sep="_") %>%
  separate(genus, c("to_delete", "genus"), sep="_") %>%
  separate(species, c("to_delete", "species"), sep="_") %>%
  dplyr::select(-to_delete)

euk.contig.bin.tax <- full_join(euk.bin.map, euk.bin.tax)
rm(euk.bin.tax, euk.bin.map)


all.contigs.tax <- rbind(contig.bin.tax,
                         euk.contig.bin.tax,
                         (contig.tax %>% mutate(bin="unbinned"))[,colnames(contig.bin.tax)],
                         (euk.tax %>% mutate(bin="unbinned"))[,colnames(contig.bin.tax)])
rm(contig.bin.tax, contig.tax,
   euk.contig.bin.tax, euk.tax)

LAO_tax <- full_join(gene.map, all.contigs.tax, by="contig")
rm(gene.map, all.contigs.tax)

```

## Data processing functions

### Filtering functions

```{r Filtering functions}

perc_filter <- function(X, perc, th, samples_arr=c()){
  # X is a count matrix whose columns are the observations (samples) and the rows are the variables (genes)
  # perc is the percentage [0-1] of observations to satisfy the condition for retaining the variable
  # th is the threshold value for the counts
  # samples_arr is a vector of sample names if subsetting is necessary
  if(length(samples_arr)==0){
    samples_arr=colnames(X)
  }
  N <- round(length(samples_arr)*perc)
  to_return <- X[rowSums(X[,samples_arr]>th)>N,]
  return(to_return)
}

censor_low <- function(X, th, rp){
  to_return <- apply(X, 2, function(x) {ifelse(x<th, rp, x)})
  return(to_return)
}

censor_low_par <- function(X, th, rp){
  cl <- makeCluster(getOption("cl.cores", 10))
  to_return <- parApply(cl=cl, X, 2, function(x) {ifelse(x<th, rp, x)})
  return(to_return)
}

test.mat <- matrix(c(0,1,2,2,4,0), nrow=2, byrow=T) %>%
  `colnames<-`(c("col1", "col2", "col3")) %>%
  `rownames<-`(c("row1", "row2"))
test.mat
censor_low(test.mat, 1.5, 0)
perc_filter(censor_low(test.mat, 1.5, 0), 0.3, 0.1)

```

### Transformation and projection

```{r Transformation and projection functions}

transform_left_basis <- function(vec, U_inv, s){
  return(U_inv %*% vec / s)
}

transform_left_basis_serial <- function(dd, U, s){
  U_inv <- ginv(U)
  dd_right <- apply(dd, 2, transform_left_basis, U_inv=U_inv, s=s)
  return(t(dd_right))
}

```

### Interpolation function

```{r Interpolation function}

get_interpolated_Sivec <- function(dd, datetime){
  for(i in datetime){
    if(!(i %in% dd$DateTime)){
      lower <- max(dd$DateTime[dd$DateTime<i])
      upper <- min(dd$DateTime[dd$DateTime>i])
      diff_time <- abs(as.numeric(min(as.numeric(difftime(lower, upper, units="mins")))))
      diff_point <- abs(as.numeric(min(as.numeric(difftime(lower, as_datetime(i), units="mins")))))
      if(diff_time<300){
        mat <- matrix(nrow=diff_time, ncol=ncol(dd)-3)
        colnames(mat) <- colnames(dd)[!colnames(dd)%in%c("Date", "Time", "DateTime")]
        mat2 <- as.data.frame(rbind(dd[dd$DateTime==lower,!colnames(dd)%in%c("Date", "Time", "DateTime")],
                      mat,
                      dd[dd$DateTime==upper,!colnames(dd)%in%c("Date", "Time", "DateTime")]))
        mat2 <- na_interpolation(mat2, option="linear")
        arr <- c(NA, NA, NA, mat2[diff_point,])
        names(arr) <- c("Date", "Time", "DateTime", colnames(dd)[!colnames(dd)%in%c("Date", "Time", "DateTime")])
        arr <- arr[colnames(dd)]
        
        dd <- rbind(dd, arr)
        dd[nrow(dd), "Date"] <- as.Date(format(as_datetime(i), format="%Y-%m-%d"))
        dd[nrow(dd), "Time"] <- (format(as_datetime(i), format="%H:%M:%S"))
        dd[nrow(dd), "DateTime"] <- as_datetime(i)
      }
    }
  }
  return(dd[dd$DateTime%in%datetime,])
}

```

### Multiplicative Replacement Strategy

Implementation of the Multiplicative Replacement Strategy from J.A. Martin-Fernández et al., Mathematical Geology, 35:3 (2003)

```{r Multiplicative Replacement Strategy}

vector_MRS <- function(x, delta){
  sum_contraint <- sum(x)
  zeros <- sum(x==0)
  multiplicative_factor <- 1-((zeros*delta)/sum_contraint)
  to_return <- ifelse(x==0, delta, x*multiplicative_factor)
  return(to_return)
}

multiplicative_replacement <- function(X, delta=NA){
  # X is a compostional matrix whose rows are the compositional vectors in agreement with compositions::clr
  if(is.na(delta)){
    delta <- 1/(ncol(X))**2 # Set default delta as 1/N**2 where N=#entries in the vectors
  }
  to_return <- apply(X, 1, vector_MRS, delta=delta)
  to_return <- to_return %>%
    `colnames<-`(rownames(X)) %>%
    `rownames<-`(colnames(X))
  return(t(to_return))
}

test.mat <- matrix(c(0,1,2,3,4,5), nrow=2, byrow=T) %>%
  `colnames<-`(c("col1", "col2", "col3")) %>%
  `rownames<-`(c("row1", "row2"))
multiplicative_replacement(test.mat)
test.noMRS <- clr(test.mat)
test.MRS <- clr(multiplicative_replacement(test.mat))
test.mat
test.noMRS
test.MRS
plot(c(test.MRS), c(test.noMRS))

```

## Data filtering

### Environmental variable collinearity reduction

```{r Reduce collinearity}

EnvEnvCorr <- cor(as.matrix(LAO_meta %>% dplyr::select(-Date, -Time, -Collector, -Aeration, -Sludge, -DateTime)), use="complete.obs")
corrplot::corrplot(EnvEnvCorr)
selected.env1 <- c("Dry_matter", "PO4.P", "Nitrat", "NH4", "Oxygen",
                  "Conductivity", "pH_manual", "Temp_manual", "Oxygen_manual")
corrplot::corrplot(cor(LAO_meta[,selected.env1], use="complete.obs"))

EnvEnvCorr <- cor(as.matrix(Sivec_meta %>% dplyr::select(-Date, -Time, -DateTime)), use="complete.obs")
corrplot::corrplot(EnvEnvCorr)
selected.env2 <- c("Inflow_Volume", "Inflow_pH", "Inflow_Conductivity",
                  "Vat1_pH", "Vat1_Sludge", "Vat1_Nitrat", "Vat1_Oxygen", "Vat1_NH4.N", "Vat1_PO4.P",
                  "Vat2_pH", "Vat2_Sludge", "Vat2_Nitrat", "Vat2_Oxygen", "Vat2_NH4.N", "Vat2_PO4.P",
                  "Areation", "Air_Pressure")
corrplot::corrplot(cor(Sivec_meta[,selected.env2], use="complete.obs"))

Sivec_interpolated <- rbind(Sivec_meta,
                            get_interpolated_Sivec(Sivec_meta,
                                                   c(LAO_meta$DateTime))) %>%
  distinct(.keep_all = T)

env_meta <- full_join(Sivec_interpolated, (LAO_meta %>% dplyr::select(-Time, -Date)))

selected.env <- c(selected.env1, selected.env2)

EnvEnvCorr <- cor(as.matrix(env_meta[, selected.env]), use="pairwise.complete.obs", method="spearman")
corrplot::corrplot(EnvEnvCorr, order = "hclust")

date2timedate <- LAO_meta$DateTime
names(date2timedate) <- LAO_meta$Date

selected.env <- c(selected.env1, selected.env2)

selected.env3 <- c("PO4.P", "Temp_manual", "Dry_matter", "Conductivity", "Areation", "Vat1_NH4.N", "Vat2_NH4.N", "Nitrat", "Inflow_Volume", "NH4", "Oxygen", "Vat2_pH", "Oxygen_manual", "pH_manual", "Inflow_pH")

```

### Check batch effect

```{r Aggregate data functions}

MG.raw[1:10000,] %>%
  pivot_longer(everything(), names_to="Sample", values_to="value") %>%
  mutate(Sample=sample_to_date[Sample]) %>%
  mutate(Group=ifelse(Sample<as.Date("2012-06-01"), "Train", "Test")) %>%
  ggplot(aes(x=value, y=factor(as.character(Sample), levels=sort(unique(as.character(Sample)), decreasing=T)), fill=Group)) +
  geom_density_ridges() +
  theme_ridges() +
  labs(x="log(Gene abundance)", y="Sample") +
  theme(axis.text.y=element_blank())

MT.raw[1:10000,] %>%
  pivot_longer(everything(), names_to="Sample", values_to="value") %>%
  mutate(Sample=sample_to_date[Sample]) %>%
  mutate(Group=ifelse(Sample<as.Date("2012-06-01"), "Train", "Test")) %>%
  ggplot(aes(x=value, y=factor(as.character(Sample), levels=sort(unique(as.character(Sample)), decreasing=T)), fill=Group)) +
  geom_density_ridges() +
  theme_ridges() +
  labs(x="log(Transcript abundance)", y="Sample") +
  theme(axis.text.y=element_blank())

MP.raw[1:10000,] %>%
  pivot_longer(everything(), names_to="Sample", values_to="value") %>%
  mutate(Sample=sample_to_date[Sample]) %>%
  mutate(Group=ifelse(Sample<as.Date("2012-06-01"), "Train", "Test")) %>%
  ggplot(aes(x=value, y=factor(as.character(Sample), levels=sort(unique(as.character(Sample)), decreasing=T)), fill=Group)) +
  geom_density_ridges() +
  theme_ridges() +
  labs(x="Protein abundance", y="Sample") +
  theme(axis.text.y=element_blank())

```


## Batch effect

### Check batch effect

```{r Aggregate data functions}

omic_type <- c("MG", "MT", "MP")

for(omic.type in omic_type){
  if(omic.type=="MP"){
    tmp.raw <- get(paste0(omic.type, ".", "raw"))
    print(paste0("Applying the percentage filter for ", omic.type, " raw data count."))
    tmp.raw <- perc_filter(tmp.raw, 0.2, 0.0000001, colnames(tmp.raw)[1:51])
    print(paste0("Applying the low value censoring for ", omic.type, " raw data count."))
    tmp.raw <- censor_low(tmp.raw, 0.0000001, 0)
    mean(colSums(get(paste0(omic.type, ".", "raw"))>0.001))
  } else {
    tmp.raw <- get(paste0(omic.type, ".", "raw"))
    print(paste0("Exp transformation the ", omic.type, " raw data count."))
    tmp.raw <- exp(tmp.raw)
    print(paste0("Applying the percentage filter for ", omic.type, " raw data count."))
    tmp.raw <- perc_filter(tmp.raw, 0.2, 0.001, colnames(tmp.raw)[1:51])
    print(paste0("Applying the low value censoring for ", omic.type, " raw data count."))
    tmp.raw <- censor_low_par(tmp.raw, 0.001, 0)
    mean(colSums(get(paste0(omic.type, ".", "raw"))>0.0000001))
  }
  tmp.raw <- multiplicative_replacement(t(tmp.raw)) # Rotate already for clr
  tmp.clr <- as.data.frame(t(clr(tmp.raw))) %>%
  `colnames<-`(sample_to_date[colnames(get(paste0(omic.type, ".", "raw")))])
  assign(paste0(omic.type, ".", "clr"), tmp.clr)
  rm(tmp.raw, tmp.clr)
}

omic_type <- c("MG", "MT")
for(omic.type in omic_type){
  
  selected.times <- sort(as_datetime(date2timedate[as.character(colnames(get(paste0(omic.type, ".", "clr"))))]))
  env.tmp <- env_meta[env_meta$DateTime%in%selected.times, c("Date", "DateTime", selected.env3)] %>%
    mutate(Date=as.Date(DateTime)) %>% arrange(Date) %>% dplyr::select(-DateTime)
  env.tmp[,selected.env3] <- as.data.frame(scale(env.tmp[,selected.env3], center=T, scale=T))
  
  selected.tech <- c("general.total_sequences", "general.avg_sequence_length")
  tech.df <- get(paste0(omic.type, ".", "multiqc")) %>%
    filter(Read_set=="R1", File_type=="preprocessed") %>%
    dplyr::select(all_of(c("metadata.sample_id", selected.tech)))
  tech.df[,selected.tech] <- as.data.frame(scale(tech.df[,selected.tech], center=T, scale=T))
  tech.df <- tech.df %>%
    mutate(Date=metadata.sample_id) %>%
    dplyr::select(-metadata.sample_id)
  
  pre.dummy <- full_join(env.tmp, tech.df) %>%
    arrange(Date) %>%
    dplyr::select(-Date)
  
  pre.dummy[is.na(pre.dummy)] <- 0
  pre.dummy <- as.matrix(pre.dummy)
  
  tmp.betas <- as.data.frame(t(lm(t(get(paste0(omic.type, ".", "clr")))~pre.dummy)$coefficients))  %>%
    `colnames<-`(c("(Intercept)", selected.env3, "general.total_sequences", "general.avg_sequence_length"))
  
  assign(paste0(omic.type, ".", "betas"), tmp.betas)
  rm(tmp.betas)
  
}

short_batch_names <- c("LAO [0/1]", "Read fail [%]", "Reads [N]", "<Read length> [Nt]", "Read dupl. [%]", "Read GC [%]")
names(short_batch_names) <- c("LAO", "general.percent_fails", "general.total_sequences", "general.avg_sequence_length","general.percent_duplicates", "general.percent_gc")

MGMT.batch.plot <- rbind((MG.betas %>%
  pivot_longer(names_to="Expl.var", values_to="beta", -"(Intercept)") %>%
  mutate(Omic="MG")),
  (MT.betas %>%
  pivot_longer(names_to="Expl.var", values_to="beta", -"(Intercept)") %>%
  mutate(Omic="MT"))) %>%
  filter(Expl.var%in%c("general.total_sequences", "general.avg_sequence_length")) %>%
  mutate(Expl.var=short_batch_names[Expl.var]) %>%
  ggplot(aes(x=Expl.var, y=beta, fill=Expl.var)) +
  geom_violin() +
  theme_classic() +
  theme(legend.position=0,
        axis.text.x=element_text(angle=60, hjust=1),
        panel.grid.major=element_line(colour="grey", size=0.25, linetype="dashed"),
        panel.grid.major.x=element_blank()) +
  facet_grid(Omic~Expl.var, scales="free") +
  labs(x="Technical variables", y="Beta distribution")

ggsave(paste0(wd, "/results/figures/MGMT_batch_paper.png"), MGMT.batch.plot, dpi=320, height=7, width=6)
ggsave(paste0(wd, "/results/figures/MGMT_batch_paper.svg"), MGMT.batch.plot, dpi=320, height=7, width=6)

```

### Further batch quantification

```{r Further batch quantifications}

omic_type <- c("MG", "MT")

shapiro.mat <- as.data.frame(matrix(nrow=length(omic_type)*4*10, ncol=4)) %>%
  `colnames<-`(c("Omic", "Expl.var", "Iteration", "P.value"))

counter <- 0
for(omic.type in omic_type){
  for(j in 1:10){
    for(k in c("general.total_sequences", "general.avg_sequence_length")){
      
      counter <- counter + 1
      shapiro.p <- shapiro.test(get(paste0(omic.type, ".", "betas"))[sample(1:nrow(get(paste0(omic.type, ".", "betas"))), 5000), k])$p.value
      shapiro.p <- formatC(shapiro.p, format = "e", digits = 1)
      shapiro.mat[counter,] <- c(omic.type, k, j, shapiro.p)
      
    }
  }
}

for(omic.type in omic_type){
  
  write_tsv(shapiro.mat %>%
              filter(Omic==omic.type) %>%
              pivot_wider(names_from=Iteration, values_from=P.value, id_cols=Expl.var), paste0(wd, "/results/tables/", omic.type,"_shapiro.tsv"))

}
  
```


### Remove technical effects

```{r Substract technical effects}

comp_tech <- function(mat.quant, mat.tech, mat.betas, index){
  as.numeric(mat.quant[index,])-(rowSums(mat.tech*t(mat.betas[index,])))
}

omic_type <- c("MG", "MT", "MP")

for(omic.type in omic_type){
  
  if(omic.type=="MP"){
    tmp.clr.corr <- get(paste0(omic.type, ".clr"))
  } else {
    
    dummy.df <- data.frame(Date=sort(sample_to_date[!names(sample_to_date)%in%c("A01", "A02")]),
                         LAO=c(rep(0, 51), rep(1, 21)))
    
    selected.tech <- c("general.total_sequences", "general.avg_sequence_length")
    tech.df <- get(paste0(omic.type, ".", "multiqc")) %>%
      filter(Read_set=="R1", File_type=="preprocessed") %>%
      dplyr::select(all_of(c("metadata.sample_id", selected.tech)))
    tech.df[,selected.tech] <- as.data.frame(scale(tech.df[,selected.tech], center=T, scale=T))
    tech.df <- tech.df %>%
      mutate(Date=metadata.sample_id) %>%
      dplyr::select(-metadata.sample_id)
    
    pre.dummy <- full_join(dummy.df, tech.df) %>%
      arrange(Date) %>%
      dplyr::select(-Date)
    
    pre.dummy[is.na(pre.dummy)] <- 0
    pre.dummy <- as.matrix(pre.dummy)
   
    a <- data.frame(pre.dummy)
    b <- get(paste0(omic.type, ".", "betas"))[,c("general.total_sequences", "general.avg_sequence_length")]
    
    cl <- makeCluster(getOption("cl.cores", 10))
    tmp.clr.corr <- as.data.frame(t(as.data.frame(parSapply(cl, as.list(1:nrow(get(paste0(omic.type, ".", "clr")))), comp_tech, mat.quant=get(paste0(omic.type, ".", "clr")), mat.tech=a, mat.betas=b)))) %>%
    `colnames<-`(c(colnames(get(paste0(omic.type, ".", "clr"))))) %>%
    `rownames<-`(c(rownames(get(paste0(omic.type, ".", "clr")))))
    
    rm(a, b, cl)
  }

  assign(paste0(omic.type, ".", "clr.corr"), tmp.clr.corr)

  rm(tmp.clr.corr)
  
}

```


## Aggragate data according to annotation

### Aggragate data functions

```{r Aggregate data functions}

reformat_ORFG_ann <- function(ORFG_names, ann, col_names){
  temp.dd <- data.frame(ORFG=ORFG_names, ORF=ORFG_names) %>%
    separate_rows(ORF, sep="\\+", convert=TRUE)
  to_return <- inner_join(temp.dd, ann[,colnames(ann)%in%c("ORF", col_names)]) %>%
    group_by(across(all_of(col_names)), ORFG) %>%
    summarise(.groups="keep")
  return(to_return)
}

```


### Aggragate data w/ functional annotation

```{r Aggregate data w/ functional annotation}

omic_type <- c("MG", "MT", "MP")
analysis_type <- c("fun")

set_type <- (expand.grid(a=omic_type, b=analysis_type) %>% mutate(d=paste(a, b, sep=".")))$d

for(i in omic_type){
  dd.tmp <- reformat_ORFG_ann(rownames(get(paste0(i, ".raw"))), LAO_ko%>%filter(A_name=="Metabolism"), c("A_name", "B_name", "C_name", "D_name"))
  dd.tmp <- inner_join(dd.tmp, get(paste0(i, ".raw")) %>% rownames_to_column(var="ORFG"))
  assign(paste0(i, ".fun"), dd.tmp)
}

KEGG_levels <- c("C_name", "D_name")
set_type.KEGG_level <- (expand.grid(a=set_type, b=KEGG_levels) %>% mutate(d=paste(a, b, sep=".")))$d

cycles.dbs <- paste(omic_type, "raw", sep=".")
counter <- 1

for(i in set_type){
  for(j in KEGG_levels){
    print(paste0(i, ".", j))
    dd.tmp <- get(i) %>%
      ungroup %>%
      dplyr::select(all_of(c(colnames(get(cycles.dbs[counter])), j)), ORFG) %>%
      pivot_longer(names_to="Sample", values_to="value", -c(all_of(j), ORFG)) %>%
      group_by(get(j), Sample) %>%
      summarise(Value=log(sum(exp(value))), .groups="keep") %>%
      ungroup() %>%
      pivot_wider(names_from=Sample, values_from=Value) %>%
      column_to_rownames(var="get(j)")
    assign(paste0(i, ".", j), dd.tmp)
  }
  counter <- counter + 1
}

```

### Aggragate data w/ taxonomic annotation

```{r Aggregate data w/ taxonomic annotation}

omic_type <- c("MG", "MT", "MP")
analysis_type <- c("tax")
tax_lvl <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")

set_type <- (expand.grid(a=omic_type, b=analysis_type) %>% mutate(d=paste(a, b, sep=".")))$d

for(i in omic_type){
  dd.tmp <- reformat_ORFG_ann(rownames(get(paste0(i, ".raw"))), LAO_tax, tax_lvl)
  dd.tmp <- inner_join(dd.tmp, get(paste0(i, ".raw")) %>% rownames_to_column(var="ORFG"))
  assign(paste0(i, ".tax"), dd.tmp)
}

set_type.tax_level <- (expand.grid(a=set_type, b=tax_lvl) %>% mutate(d=paste(a, b, sep=".")))$d

cycles.dbs <- paste(omic_type, "raw", sep=".")
counter <- 1

for(i in set_type){
  for(j in rev(tax_lvl)){
    print(paste0(i, ".", j))

    dd.tmp <- get(i) %>%
    replace_na(list(superkingdom="not classified", phylum="not classified",
                    class="not classified", order="not classified",
                    family="not classified", genus="not classified",
                    species="not classified")) %>%
    ungroup %>%
    dplyr::select(all_of(c(colnames(get(i)), tax_lvl)), ORFG) %>%
    pivot_longer(names_to="Sample", values_to="value", -c(all_of(tax_lvl), ORFG)) %>%
    `colnames<-`(c(tax_lvl, "ORFG", "Sample", "value"))

    if(i=="MP.tax"){
      dd.tmp <- dd.tmp %>% group_by(get(j), Sample) %>%
      summarise(Value=sum(value), .groups="keep") %>%
      ungroup()
    } else {
      dd.tmp <- dd.tmp %>% group_by(get(j), Sample) %>%
      summarise(Value=log(sum(exp(value))), .groups="keep") %>%
      ungroup()
    }
    
    dd.tmp <- dd.tmp %>%
      pivot_wider(names_from=Sample, values_from=Value) %>%
      column_to_rownames(var="get(j)")
    assign(paste0(i, ".", j), dd.tmp[,colnames(get(i))[9:ncol(get(i))]])
  }
  
  counter <- counter + 1
}

```

## Split train/test data

### Split data functions

```{r split data}

prep_data_mats <- function(mat, multiqc_table){
  
  mat <- exp(mat)
  mat <- perc_filter(mat, 0.2, 0.001, colnames(mat)[1:51])
  mat <- censor_low(mat, 0.001, 0)
  mat <- as.data.frame(t(clr(multiplicative_replacement(t(mat)))))
  mat <- as.data.frame(t(clr(t(mat))))
  colnames(mat) <- sample_to_date[colnames(mat)]
  
  selected.times <- sort(as_datetime(date2timedate[as.character(colnames(get(paste0(omic.type, ".", "clr"))))]))
  env.tmp <- env_meta[env_meta$DateTime%in%selected.times, c("Date", "DateTime", selected.env3)] %>%
    mutate(Date=as.Date(DateTime)) %>% arrange(Date) %>% dplyr::select(-DateTime)
  env.tmp[,selected.env3] <- as.data.frame(scale(env.tmp[,selected.env3], center=T, scale=T))
  
  selected.tech <- c("general.total_sequences", "general.avg_sequence_length")
  tech.df <- get(paste0(omic.type, ".", "multiqc")) %>%
    filter(Read_set=="R1", File_type=="preprocessed") %>%
    dplyr::select(all_of(c("metadata.sample_id", selected.tech)))
  tech.df[,selected.tech] <- as.data.frame(scale(tech.df[,selected.tech], center=T, scale=T))
  tech.df <- tech.df %>%
    mutate(Date=metadata.sample_id) %>%
    dplyr::select(-metadata.sample_id)

  pre.dummy <- full_join(env.tmp, tech.df) %>%
    arrange(Date) %>%
    dplyr::select(-Date)
  
  pre.dummy[is.na(pre.dummy)] <- 0
  pre.dummy <- as.matrix(pre.dummy)
  
  tmp.betas <- as.data.frame(t(lm(t(get(paste0(omic.type, ".", "clr")))~pre.dummy)$coefficients))  %>%
    `colnames<-`(c("(Intercept)", selected.env3, "general.total_sequences", "general.avg_sequence_length"))
    
    
  a <- as.data.frame(pre.dummy)
  b <- tmp.betas[,c("general.total_sequences", "general.avg_sequence_length")]
  
  cl <- makeCluster(getOption("cl.cores", 10))
  mat.clr.corr <- as.data.frame(t(as.data.frame(parSapply(cl, as.list(1:nrow(mat)), comp_tech, mat.quant=mat, mat.tech=a, mat.betas=b)))) %>%
  `colnames<-`(c(colnames(mat))) %>%
  `rownames<-`(c(rownames(mat)))
  
  mat.train.clr.corr <- mat.clr.corr[, colnames(mat.clr.corr) < "2012-05-31"]
  mat.train.svd.corr <- svd(mat.train.clr.corr)
  
  mat.test.clr.corr <- mat.clr.corr[, colnames(mat.clr.corr) > "2012-05-31"]
  mat.test.svd.corr <- transform_left_basis_serial(mat.test.clr.corr, mat.train.svd.corr$u, mat.train.svd.corr$d)
  
  return(list(mat.train.clr.corr, mat.test.clr.corr, mat.train.svd.corr, mat.test.svd.corr))
  
}

prep_data_mats_var <- function(mat){
  
  mat <- perc_filter(mat, 0.2, 0.0000001, colnames(mat)[1:51])
  mat <- censor_low(mat, 0.0000001, 0)
  
  mat <- as.data.frame(t(clr(multiplicative_replacement(t(mat)))))
  colnames(mat) <- sample_to_date[colnames(mat)]
  
  mat.train.clr.corr <- mat
  mat.train.svd.corr <- svd(mat.train.clr.corr)
  
  mat.test.clr.corr <- ""
  mat.test.svd.corr <- ""
  
  return(list(mat.train.clr.corr, mat.test.clr.corr, mat.train.svd.corr, mat.test.svd.corr))
  
}

```


### Quantitative exploration

```{r Quantitative exploration}

hist(apply(MT.tax.genus, 1, quantile, probs=c(.75)))

hist(apply(MG.fun.C_name, 1, quantile, probs=c(.75)))

MG.fun.C_name[apply(MG.fun.C_name, 1, quantile, probs=c(.75))>11,] %>%
  rownames_to_column("Reaction") %>%
  pivot_longer(names_to="Sample", values_to="value", -Reaction) %>%
  mutate(Sample=sample_to_date[Sample]) %>%
  ggplot(aes(x=Sample, y=value, fill=Reaction)) +
  geom_bar(stat="identity")

a <- t.test(exp(as.numeric(MG.fun.C_name["Fatty acid degradation", 11:14])), exp(as.numeric(MG.fun.C_name["Fatty acid degradation", 52:56])))

plot(log(c(exp(as.numeric(MG.fun.C_name["Fatty acid degradation", 1:51])), exp(as.numeric(MG.fun.C_name["Fatty acid degradation", 52:72]))-(a$estimate[2]-a$estimate[1]))), type="b")

```


### Data split and correct

```{r Data split}

omic_type <- c("MG", "MT", "MP")
#omic_type <- c("MP")
fun_lvl <- paste("fun", c("C_name", "D_name"), sep=".")
tax_lvl <- paste("tax", c("superkingdom", "phylum", "class", "order", "family", "genus", "species"), sep=".")
subsets <- c(fun_lvl, tax_lvl)
#set_type <- c((expand.grid(a=omic_type, b=subsets) %>% mutate(d=paste(a, b, sep=".")))$d)

for(omic.type in omic_type){
  for(i in subsets){
    j <- paste0(omic.type, ".", i)
    print(j)
    if(omic.type=="MP"){
      cycle_list <- prep_data_mats_var(get(j)[,names(sort(sample_to_date[colnames(get(j))]))]) # sort?
    } else {
      cycle_list <- prep_data_mats(get(j)[,names(sort(sample_to_date[colnames(get(j))]))], MG.multiqc) # sort?
    }
    assign(paste0(j, ".", "train.clr.corr"), cycle_list[[1]])
    assign(paste0(j, ".", "test.clr.corr"), cycle_list[[2]])
    assign(paste0(j, ".", "train.svd.corr"), cycle_list[[3]])
    assign(paste0(j, ".", "test.svd.corr"), cycle_list[[4]])
  }
      
  assign(paste0(omic.type, ".raw.train.clr.corr"),
         get(paste0(omic.type, ".clr.corr"))[,sort(colnames(get(paste0(omic.type, ".clr.corr"))))][, colnames(get(paste0(omic.type, ".clr.corr"))) < "2012-05-31"])
  assign(paste0(omic.type, ".raw.train.svd.corr"),
         svd(get(paste0(omic.type, ".raw.train.clr.corr"))))
    
  assign(paste0(omic.type, ".raw.test.clr.corr"),
         get(paste0(omic.type, ".clr.corr"))[sort(colnames(get(paste0(omic.type, ".clr.corr"))))][, colnames(get(paste0(omic.type, ".clr.corr"))) > "2012-05-31"])
  assign(paste0(omic.type, ".raw.test.svd.corr"),
         transform_left_basis_serial(get(paste0(omic.type, ".raw.test.clr.corr")),
                                     get(paste0(omic.type, ".raw.train.svd.corr"))$u,
                                     get(paste0(omic.type, ".raw.train.svd.corr"))$d))
}

```


### Print first EGs

```{r Print first EGs}

for(i in 1:5){
  simple.plot <- data.frame(Sample=as.Date(colnames(MG.raw.train.clr.corr)), value=MG.raw.train.svd.corr$v[,i]) %>%
    ggplot(aes(x=Sample, y=value, group=1)) +
    geom_point() +
    geom_line() +
    theme_minimal()
  print(simple.plot)
}
for(i in 1:5){
  simple.plot <- data.frame(Sample=as.Date(colnames(MT.raw.train.clr.corr)), value=MT.raw.train.svd.corr$v[,i]) %>%
    ggplot(aes(x=Sample, y=value, group=1)) +
    geom_point() +
    geom_line() +
    theme_minimal()
  print(simple.plot)
}

for(i in 1:5){
  simple.plot <- data.frame(Sample=as.Date(colnames(MG.fun.C_name.train.clr.corr)), value=as.numeric(MG.fun.C_name.train.clr.corr[i,])) %>%
    ggplot(aes(x=Sample, y=value, group=1)) +
    geom_point() +
    geom_line() +
    theme_minimal()
  print(simple.plot)
}

for(i in 1:5){
  simple.plot <- data.frame(Sample=as.Date(colnames(MG.fun.C_name.train.clr.corr)), value=as.numeric(MG.fun.C_name.train.svd.corr$v[,i])) %>%
    ggplot(aes(x=Sample, y=value, group=1)) +
    geom_point() +
    geom_line() +
    theme_minimal()
  print(simple.plot)
}

```

### Experimantal batches

```{r Experimantal batches}

date_to_sample <- names(sample_to_date)
names(date_to_sample) <- sample_to_date
tmp.mat <- MG.raw.train.svd.corr$v
plot(tmp.mat[,1], type = "b")
rownames(tmp.mat) <- date_to_sample[colnames(MG.raw.train.clr.corr)]
tmp.mat <- tmp.mat[sort(rownames(tmp.mat)),]
plot(tmp.mat[,1], type = "b")

```


### Output split data

```{r Output split data}

omic_type <- c("MG", "MT", "MP")
fun_lvl <- paste("fun", c("C_name", "D_name"), sep=".")
tax_lvl <- paste("tax", c("superkingdom", "phylum", "class", "order", "family", "genus", "species"), sep=".")
subsets <- c(fun_lvl, tax_lvl)
set_type <- c((expand.grid(a=omic_type, b=subsets) %>% mutate(d=paste(a, b, sep=".")))$d)

save(file=paste0(wd, "/preprocessed/preprocessed_mats_MRS_sampels.RData"),
     list=c((expand.grid(a=c(set_type, "MG.raw", "MT.raw", "MP.raw"),
                         b=c(".train.clr.corr", ".test.clr.corr", ".train.svd.corr", ".test.svd.corr"))
             %>% mutate(d=paste0(a, b)))$d), set_type)

```

